Random number generation and applications on GPU’s in R?

### 2021-09-17 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
add paper1 to inst
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2851 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
updates paper1 in inst folder
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2855 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

We show the usage of \fct{gpuRandom::rnorm} by generating a $512 \times 1024$ matrix named \code{normal_matrix} of double presicion Gaussian random numbers, and assess Normality by making a histogram (left panel of Figure \ref{fig2}) and a Q-Q plot (right panel of Figure \ref{fig2}) of the generated random numbers. The points in the Q-Q plot in Figure \ref{fig2} fall along the diagonal line, we have good evidence the data are from Gaussian distribution. We find doing a Q-Q plot can occupy a large amount of time in the process of producing an R markdown file, so as an incidental work, we make the function \fct{gpuRandom::qqnorm} that does the quantile computation for the normal distribution part (adapted from \fct{stat::qnorm}) on GPU, which speeds up the Q-Q plot greatly.
<<histAndQQ_prepare,eval=TRUE>>=
streams <- CreateStreamsGpu(Nstreams =512*128, keepInitial=1)
normal_matrix<-gpuRandom::rnorm(c(1024,512), streams=streams, 
                                Nglobal=c(512,128), type="double")
@
<<histAndQQ_prepare_for_cache, eval=TRUE, cache=TRUE>>=
avector<-as.vector(as.matrix(normal_matrix))
@
\begin{figure}[H]
\centering
 <<histAndQQ, echo=TRUE, eval=TRUE, cache=TRUE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(avector,breaks=40, main = "histogram")
gpuRandom::qqnorm(avector, Nglobal=c(128,64), main="Q-Q Plot")
@
\caption{\label{fig2} Histogram and Q-Q plot of the GPU-generated random numbers.}
\end{figure}
Let's compare the time of generating a large-size matrix of more than 26.8 million Gaussian random numbers of double precision using the \fct{stats::rnorm} and using the \fct{gpuRandom::rnorm}. Both process are not time-consuming, however, using \fct{gpuRandom::rnorm} with $128 \times 128$ work-items is about 10 times (the number may fluctuate a bit from executions) faster than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
<<random normal time compare, eval=TRUE, cache=FALSE>>=
system.time(gpuRandom::rnorm(c(16384,16384), streams=streams, 
                             Nglobal=c(512,128), type="double"))
system.time(matrix(stats::rnorm(16384*16384),16384,16384))
@

@@

Next we compare the run time between using \fct{gpuRandom::fisher.sim} and \fct{stats::fisher.test} using two real data examples with small and large p-values respectively. These 2-way contingency tables (\ref{tab:month} and \ref{tab:week}) consists of selected data from the 2018 Natality public use file \citep{National2018} from the Centers for Disease Control and Prevention’s National Center for Health Statistics, the 2018 natality data file may be downloaded at \url{https://www.cdc.gov/nchs/data_access/VitalStatsOnline.htm}. The row variable of the first table represents the twelve months: from January to December, and for the second table it represents the seven weekdays: Monday to Sunday. The column variables of these two tables represent the twelve categories of congenital anomalies of the newborn: 1) Anencephaly; 2) Meningomyelocele/Spina bifida; 3) Cyanotic congenital heart disease; 4) Congenital diaphragmatic hernia; 5) Omphalocele; 6) Gastrochisis; 7) Limb reduction defect; 8) Cleft lip with or without cleft palate; 9) Cleft palate alone; 10) Down syndrome; 11) Suspected chromosomal disorder; and 12) Hypospadias.

\subsection{Comparing running time: Month data example}
<<Timecomparemonth,echo=FALSE,eval=TRUE, results="hide">>=
data2<-read.csv("/home/ruoyong/paper2/month.csv") 
data2<-data2[,-1]
month<-as.matrix(data2)
rownames(month) <- c("Jan", "Feb","Mar","Apr","May","Jun","Jul","Aug","Sept","Oct","Nov","Dec")
colnames(month) <- c("Ane", "Men", "Cya", "Her", "Omp", "Gas", "Lim", "Cle", "Pal", "Down", "Chro", "Hypo")
Nglobal = c(128,128)
streams <- CreateStreamsGpu(Nstreams =128*128, keepInitial=1)
@
<<monthdata,eval=TRUE,echo=FALSE, cache=TRUE>>=
library(knitr)
library(kableExtra)
knitr::kable(head(month, n=12),format="latex",  caption = "Month\\label{tab:month}") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center",latex_options = "HOLD_position")
@
<<timecomparemonth2Cpu, cache=TRUE, eval=TRUE>>=
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1e6)$p.value
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1e6))
@
\fct{stats::fisher.test} takes about 9.8 seconds (time varies a little bit from executions) for one million simulations, has a p-value of 0.4035.
<<timecomparemonth2Gpu, eval=TRUE,cache=FALSE>>=
#using GPU
almost.1 <- 1 + 64 * .Machine$double.eps
observed_m= -gpuRandom::logfactSum(month, c(16,16))/almost.1
month_GPU<-vclMatrix(month,type="integer")
system.time(gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = Nglobal))
@
\fct{gpuRandom::fisher.sim} takes about 2.3 seconds, compared to the elapsed time of \fct{gpuRandom::fisher.sim},  \fct{gpuRandom::fisher.sim} takes less than a quater of the time. 
<<timecomparemonth3Gpu, eval=TRUE,cache=TRUE>>=
result_monthgpu <- gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = Nglobal)
result_monthgpu$simnum
result_monthgpu$counts
result_monthgpu$p.value
@
We requested one million simulations on GPU while the actual number of simulation is 999424, and got 403274 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.4035, which is close enough to the p-value from \fct{stats::fisher.test}: 0.4035.  
<<timecomparemonth4Gpu, eval=TRUE,echo=FALSE, cache=TRUE>>=
result_monthgpu$cpu <- as.vector(result_monthgpu$sim)
@

\subsection{Comparing running time: Week data example}
<<Timecompareweek, echo=FALSE, eval=TRUE, results="hide">>=
library(latex2exp)
data<-read.csv("/home/ruoyong/paper2/weekday.csv") 
#exampletwo<-as.matrix(as.data.frame(lapply(data, as.numeric)))
data<-data[,-1]
week<-as.matrix(data)
rownames(week) <- c("Mon", "Tue","Wed","Thu","Fri","Sat","Sun")
colnames(week) <- c("Ane", "Men", "Cya", "Her", "Omp", "Gas", "Lim", "Cle", "Pal", "Down", "Chro", "Hypo")
@

<<weekdata, eval=TRUE, echo=FALSE, cache=TRUE>>=
knitr::kable(head(week, n=7),format="latex",caption = "Week\\label{tab:week}") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center",latex_options = "HOLD_position")
@

<<Timecompareweek2Cpu, cache=TRUE, eval=TRUE>>=
#using CPU
fisher.test(week,simulate.p.value = TRUE,B=1e7)$p.value
system.time(fisher.test(week,simulate.p.value = TRUE,B=1e7))
@
The ``week'' table has a much smaller p-value: around 0.000119, which should require a larger number of simulations to get a more accurate p-value.
<<Timecompareweek2Gpu, eval=TRUE, cache=FALSE>>=
#using GPU
observed_w= -gpuRandom::logfactSum(week, c(16,16))/almost.1
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
            type="double", returnStatistics=FALSE,Nglobal = Nglobal))
@
\fct{stats::fisher.test} takes 60.4 seconds, while \fct{fisher.sim} takes about 10.8 seconds, the elapsed time is decreased to about 20\% of the time taken on CPU. 
<<Timecompareweek2GpuTest, eval=TRUE, cache=TRUE>>=
result_weekgpu<-gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = Nglobal)
result_weekgpu$simnum
result_weekgpu$counts
result_weekgpu$p.value
@
With about ten million simulations, we get 1272 cases and a p-value around 0.0001273. We summarized the results in Table \ref{tab:summary}.
<<Timecompareweek3GpuTest, echo=FALSE, eval=TRUE, cache=TRUE>>=
result_weekgpu$cpu = as.vector(result_weekgpu$sim)
@
\begin{table}[H]
\begin{tabular}{ |c|c|c|c|c|c| } 
 \hline
 test example & B & p-value stats & p-value gpuRandom & time stats & time gpuRandom \\ \hline
 Month & 1e6 & 0.4029 & 0.4035 & 9.8 & 2.3\\ \hline
 Week & 1e7 & 0.000119 & 0.000127 & 60.4 & 10.8\\ \hline
\end{tabular}
\caption{\label{tab:summary} Comparison summary of \fct{gpuRandom::fisher.sim} and \fct{stats::fisher.test}.}
\end{table}

@@

\subsection{Simulating Gaussian random fields with Mat\'ern covariances}
<<setupData, eval=TRUE,echo=FALSE, message=FALSE>>=
library("gpuR")
library("gpuRandom")
library('geostatsp')
@
Step 1, set up a $60 \times 80$ raster and coordinates on the raster, convert the matrix of coordinates \code{coordsSp@coords} to a ``vclMatrix'' object for later use.
<<setupcoords, eval=TRUE>>=
Ngrid = c(60, 80)
NlocalCache = 1000
Nglobal = c(128, 64, 2)
Nlocal = c(4, 2, 2)
theType = "float"
myRaster = raster(
  extent(0,Ngrid[1]/Ngrid[2],5,6),
  Ngrid[2], Ngrid[1]
)
coordsSp = sp::SpatialPoints(xyFromCell(myRaster, 1:ncell(myRaster)))
coordsGpu = vclMatrix(coordsSp@coords, 
                      nrow(coordsSp@coords), 
                      ncol(coordsSp@coords), type=theType)
@
Step 2, create 5 parameter sets as a small example, in practical studies, there can be hundreds or thousands of parameter sets. Then, convert the matrix of parameters to a ``vclMatrix'' for later use by \fct{gpuRandom::maternGpuParam}.
<<setup paramsBatch, eval=TRUE,echo=FALSE>>=
myParamsBatch = 
rbind(c(shape=1.25, range=0.5, variance = 1.5, nugget = 0,anisoRatio = 1, anisoAngleRadians = 0), 
c(shape=2.15, range=0.25, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=0.55, range=1.5, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=2.15, range=0.5, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = -pi/7),
c(shape=2.15, range=0.5, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/4))
@
<<show parameterBatch,eval=TRUE>>=
myParamsBatch
paramsGpu = gpuRandom::maternGpuParam(myParamsBatch, type=theType)
@
Step 3, compute Mat\'ern covariance matrices using \fct{gpuRandom::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4800 \times 4800$ and are stacked by row in the output \code{maternCov}. 
<<r outputBatchCreate, eval=TRUE>>=
maternCov = vclMatrix(0, nrow(paramsGpu)*nrow(coordsGpu), 
                         nrow(coordsGpu),type=theType)
dim(maternCov)
gpuRandom::maternBatch(maternCov, coordsGpu, paramsGpu,
                       Nglobal=c(64,64,16), Nlocal=c(4,4,4))
@
Step 4, the first argument \code{maternCov} in \fct{gpuRandom::cholBatch} specifies the matrix object to take Cholesky decomposition. Unit lower triangular matrices $L_i$'s are returned and stacked by row in \code{maternCov}, \code{diagMat} stores the diagonal values of each $D_i$ in a row, for example, if each batch $\Sigma_i$ is of size $n \times n$, then each batch $L_i$ is $n \times n$, and each batch $D_i$ is $1 \times n$.
<<r cholBach, eval=TRUE>>=
diagMat = vclMatrix(0, nrow(paramsGpu), ncol(maternCov), type = theType)
gpuRandom::cholBatch(maternCov, diagMat, numbatchD=nrow(myParamsBatch),
                     Nglobal,Nlocal,NlocalCache)
@
% \begin{gather}
% \begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
% \end{bmatrix}
%  \rightarrow
%  \begin{bmatrix}
%   L_{1} \\ L_{2} \\L_{3} \\ \vdots
%   \end{bmatrix} \text{and}
%   \begin{bmatrix}
%   D_{1} \\ D_{2} \\D_{3} \\ \vdots
%   \end{bmatrix}
% \end{gather}
Step 5, generate 2 standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{gpuRandom::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
<<r randomNormalsSim1, eval=TRUE, cache=FALSE>>=
streamsGpu <- gpuRandom::CreateStreamsGpu(Nstreams=128*64, keepInitial=TRUE)
@
<<r randomNormalsSim2, eval=TRUE, cache=FALSE>>=
zmatGpu = gpuRandom::rnorm(c(nrow(maternCov),2), 
                           streams=streamsGpu, Nglobal=c(128,64),
                           type = theType)
@
Step 6, use \fct{gpuRandom::multiplyLowerDiagonalBatch} to compute $U = L * D^{(1/2)}* Z$ in batches, see the following illustration. \code{simMat} is the output matrix for $U$, \code{maternCov}, \code{diagMat}, and \code{zmatGpu} correspond to the matrices of $L$, $D$ and $Z=(Z_1, Z_2)$ respectively.
<<r multLowerdiag_LDZ, eval=TRUE, cache=FALSE, results='hide'>>=
simMat = vclMatrix(0, nrow(zmatGpu), ncol(zmatGpu), 
                   type = theType)

gpuRandom::multiplyLowerDiagonalBatch(
  simMat, maternCov, diagMat, zmatGpu,
  diagIsOne = 1L, # diagonal of L is one
  transformD = "sqrt", # take the square root of each element of D
  Nglobal,
  Nlocal,
  NlocalCache)
@
\begin{gather}
 \begin{bmatrix}  L_{1} \\ L_{2} \\L_{3} \\ \vdots
 \end{bmatrix} 
 *
  \begin{bmatrix}
   D_{1} \\ D_{2} \\D_{3} \\ \vdots
   \end{bmatrix} 
   *
   \begin{bmatrix}
   Z_{11} & Z_{12}
   \end{bmatrix}
  =
 \begin{bmatrix}
   L_1D_1Z_{11} & L_1D_1Z_{12} \\
   L_2D_2Z_{11} & L_2D_2Z_{12} \\
   L_3D_3Z_{11} & L_3D_3Z_{12} 
  \end{bmatrix}
\end{gather}
Step 7, Finally, plot the 10 simulated Gaussian random surfaces.
<<r plotSimSetup, echo=TRUE, eval=TRUE, cache=TRUE>>=
library(raster)
simRaster = brick(myRaster, nl = ncol(simMat)*nrow(paramsGpu))
values(simRaster) = as.vector(as.matrix(simMat))
names(simRaster) = apply(expand.grid('par',1:nrow(paramsGpu), 
                                     'sim', 1:ncol(simMat)), 1, paste, collapse='')
@
\begin{figure}[H]
\centering
<<maternplot, echo=TRUE, fig.show='hold', eval=TRUE, dev='pdf',cache=TRUE>>=
par(mar=rep(0.1, 4))
for(D in names(simRaster)) {
  plot(extent(simRaster))
  plot(simRaster[[D]], legend=FALSE, add=TRUE)
}
@
\caption{\label{fig5} }
\end{figure}

>>>>

We show the usage of \fct{gpuRandom::rnorm} by generating a $512 \times 1024$ matrix named \code{normal_matrix} of double presicion Gaussian random numbers, and assess Normality by making a histogram (left panel of Figure \ref{fig2}) and a Q-Q plot (right panel of Figure \ref{fig2}) of the generated random numbers. The points in the Q-Q plot in Figure \ref{fig2} fall along the diagonal line, we have good evidence the data are from Gaussian distribution. We find doing a Q-Q plot can occupy a large amount of time in the process of producing an R markdown file, so as an incidental work, we make the function \fct{gpuRandom::qqnorm} that does the quantile computation for the normal distribution part (adapted from \fct{stat::qnorm}) on GPU, which speeds up the Q-Q plot greatly.
<<histAndQQ_prepare,eval=TRUE>>=
streams <- CreateStreamsGpu(Nstreams =512*128, keepInitial=1)
normal_matrix<-gpuRandom::rnorm(c(1024,512), streams=streams, 
                                Nglobal=c(512,128), type="double")
@
<<histAndQQ_prepare_for_cache, eval=TRUE, cache=TRUE>>=
avector<-as.vector(as.matrix(normal_matrix))
@
\begin{figure}[H]
\centering
 <<histAndQQ, echo=TRUE, eval=TRUE, cache=TRUE, fig.height=4, fig.width=4, fig.show='hold'>>=
hist(avector,breaks=40, main = "histogram")
gpuRandom::qqnorm(avector, Nglobal=c(128,64), main="Q-Q Plot")
@
\caption{\label{fig2} Histogram and Q-Q plot of the GPU-generated random numbers.}
\end{figure}
Let's compare the time of generating a large-size matrix of 100 million Gaussian random numbers of double precision using the \fct{stats::rnorm} and using the \fct{gpuRandom::rnorm}. Both process are not time-consuming, however, using \fct{gpuRandom::rnorm} with $512 \times 128$ work-items is 117.6 times (the number may fluctuate a bit from executions) faster than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
<<random normal time compare, eval=TRUE, cache=TRUE>>=
system.time(gpuRandom::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
system.time(matrix(stats::rnorm(10000^2),10000,10000))
@

@@

We show the usage and advantage of \fct{gpuRandom::fisher.sim} by computing the p-values for two real data real data examples: one with a relatively big p-value and another with a very small p-value, and compare the run time with using \fct{stats::fisher.test} for each of the data sets on two computers: one with a very good CPU and a normal GPU, the other is equipped with an excellent GPU and a normal CPU. The device information is included in the table \ref{tab:device}. The \proglang{R} outputs for testing on computer 1 is shown in the following sections. 
\begin{table}[H]
\begin{tabular}{ |c|c|c|c|c|c| } 
\hline
           & CPU   & GPU \\ \hline
 Computer1 & pthread-Intel(R) Xeon(R) W-2145 CPU @ 3.70GHz & 2 Radeon VII's : ROCm AMDGPU  \\ \hline
 Computer2 & dontknow & GRID V100D-8C \\ \hline
\end{tabular}
\caption{\label{tab:device} Device information for run-time compare.}
\end{table}

These 2-way contingency tables (\ref{tab:month} and \ref{tab:week}) consists of selected data from the 2018 Natality public use file \citep{National2018} from the Centers for Disease Control and Prevention’s National Center for Health Statistics, the 2018 natality data file may be downloaded at \url{https://www.cdc.gov/nchs/data_access/VitalStatsOnline.htm}. The row variable of the first table represents the twelve months: from January to December, and for the second table it represents the seven weekdays: Monday to Sunday. The column variables of these two tables represent the twelve categories of congenital anomalies of the newborn: 1) Anencephaly; 2) Meningomyelocele/Spina bifida; 3) Cyanotic congenital heart disease; 4) Congenital diaphragmatic hernia; 5) Omphalocele; 6) Gastrochisis; 7) Limb reduction defect; 8) Cleft lip with or without cleft palate; 9) Cleft palate alone; 10) Down syndrome; 11) Suspected chromosomal disorder; and 12) Hypospadias.

\subsection{Comparing running time: Month data example}
<<Timecomparemonth,echo=FALSE,eval=TRUE, results="hide">>=
data2<-read.csv("/home/ruoyong/paper1_2021_5/month.csv") 
data2<-data2[,-1]
month<-as.matrix(data2)
rownames(month) <- c("Jan", "Feb","Mar","Apr","May","Jun","Jul","Aug","Sept","Oct","Nov","Dec")
colnames(month) <- c("Ane", "Men", "Cya", "Her", "Omp", "Gas", "Lim", "Cle", "Pal", "Down", "Chro", "Hypo")
Nglobal = c(128,128)
streams <- CreateStreamsGpu(Nstreams =128*128, keepInitial=1)
@
<<monthdata,eval=TRUE,echo=FALSE, cache=TRUE>>=
library(knitr)
library(kableExtra)
knitr::kable(head(month, n=12),format="latex",  caption = "Month\\label{tab:month}") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center",latex_options = "HOLD_position")
@
<<timecomparemonth2Cpu, cache=TRUE, eval=TRUE>>=
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1e6)$p.value
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1e6))
@
\fct{stats::fisher.test} takes about 9.8 seconds (time varies a little bit from executions) for one million simulations, has a p-value of 0.4035.
<<timecomparemonth2Gpu, eval=TRUE,cache=FALSE>>=
#using GPU
almost.1 <- 1 + 64 * .Machine$double.eps
observed_m= -gpuRandom::logfactSum(month, c(16,16))/almost.1
month_GPU<-vclMatrix(month,type="integer")
system.time(gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = c(128,128)))
@
\fct{gpuRandom::fisher.sim} takes about 2.3 seconds, compared to the elapsed time of \fct{gpuRandom::fisher.sim},  \fct{gpuRandom::fisher.sim} takes less than a quater of the time. 
<<timecomparemonth3Gpu, eval=TRUE,cache=FALSE>>=
result_monthgpu <- gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(128,128))
result_monthgpu$simnum
result_monthgpu$counts
result_monthgpu$p.value
@
We requested one million simulations on GPU while the actual number of simulation is 999424, and got 403274 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.4035, which is close enough to the p-value from \fct{stats::fisher.test}: 0.4034.  
<<timecomparemonth4Gpu, eval=TRUE,echo=FALSE, cache=TRUE>>=
result_monthgpu$cpu <- as.vector(result_monthgpu$sim)
@

\subsection{Comparing running time: Week data example}
<<Timecompareweek, echo=FALSE, eval=TRUE, results="hide">>=
library(latex2exp)
data<-read.csv("/home/ruoyong/paper1_2021_5/weekday.csv") 
#exampletwo<-as.matrix(as.data.frame(lapply(data, as.numeric)))
data<-data[,-1]
week<-as.matrix(data)
rownames(week) <- c("Mon", "Tue","Wed","Thu","Fri","Sat","Sun")
colnames(week) <- c("Ane", "Men", "Cya", "Her", "Omp", "Gas", "Lim", "Cle", "Pal", "Down", "Chro", "Hypo")
@

<<weekdata, eval=TRUE, echo=FALSE, cache=TRUE>>=
knitr::kable(head(week, n=7),format="latex",caption = "Week\\label{tab:week}") %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "center",latex_options = "HOLD_position")
@
We perform 10 million simulations for the ``week'' table.
<<TimecompareweekCpu, cache=TRUE, eval=TRUE>>=
#using CPU
fisher.test(week,simulate.p.value = TRUE,B=1e7)$p.value
system.time(fisher.test(week,simulate.p.value = TRUE,B=1e7))
@
The ``week'' table has a much smaller p-value: around 0.000119, which should require a larger number of simulations to get a more accurate p-value.
<<TimecompareweekGpu, eval=TRUE, cache=FALSE>>=
#using GPU
observed_w= -gpuRandom::logfactSum(week, c(16,16))/almost.1
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
            type="double", returnStatistics=FALSE,Nglobal = Nglobal))
@
\fct{stats::fisher.test} takes 60.4 seconds, while \fct{fisher.sim} takes about 10.8 seconds, the elapsed time is decreased to about 20\% of the time taken on CPU. 
<<TimecompareweekGpu2, eval=TRUE, cache=FALSE>>=
result_weekgpu<-gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = Nglobal)
result_weekgpu$simnum
result_weekgpu$counts
result_weekgpu$p.value
@
With about ten million simulations, we get 1272 cases and a p-value around 0.0001274. We summarized the results in Table \ref{tab:summary}.
<<TimecompareweekGpu3, echo=FALSE, eval=TRUE, cache=TRUE>>=
result_weekgpu$cpu = as.vector(result_weekgpu$sim)
@
\begin{table}[H]
\centering
\begin{tabular}{ |c|c|c|c|c|c| } 
 \hline
 data &   B  & Intel CPU &  AMD GPU & NVIDIA GPU \\ \hline
 Month & 1e6 & 0.4038 & 0.403507  & 0.403507     \\ \hline
 Week &  1e7  & 0.0001208 & 0.0001274 & 0.0001274\\ \hline
\end{tabular}
\caption{\label{tab:summary} Summary of p-values obtained on different devices.}
\end{table}

\begin{table}[H]
\centering
\begin{tabular}{ |c|c|c|c|c|c| } 
 \hline
 data &   B  & Intel CPU & sencha CPU & AMD GPU & NVIDIA GPU \\ \hline
 Month & 1e6 & 9.74 & 14.58 & 2.28 & 0.31\\ \hline
 Week &  1e7  & 60.19 & 90.19 & 10.82 & 1.90\\ \hline
\end{tabular}
\caption{\label{tab:summary} Run-time compare of Fisher's test simulation on different devices, all numbers are in seconds.}
\end{table}

@@

\subsection{Simulating Gaussian random fields with Mat\'ern covariances}
<<setupData, eval=TRUE,echo=FALSE, message=FALSE>>=
library("gpuR")
library("gpuRandom")
library('geostatsp')
@
Step 1, set up a $60 \times 80$ raster and coordinates on the raster, convert the matrix of coordinates \code{coordsSp@coords} to a ``vclMatrix'' object for later use.
<<setupcoords, eval=TRUE>>=
Ngrid = c(60, 80)
NlocalCache = 1000
Nglobal = c(128, 64, 2)
Nlocal = c(4, 2, 2)
theType = "double"

myRaster = raster::raster( raster::extent(0,Ngrid[1]/Ngrid[2],5,6),Ngrid[1], Ngrid[2])
coordsSp = sp::SpatialPoints(raster::xyFromCell(myRaster, 1:raster::ncell(myRaster)))
coordsGpu = vclMatrix(coordsSp@coords, 
                      nrow(coordsSp@coords), 
                      ncol(coordsSp@coords), type=theType)
@
Step 2, create 5 parameter sets as a small example, in practical studies, there can be hundreds or thousands of parameter sets. Then, convert the matrix of parameters to a ``vclMatrix'' for later use by \fct{gpuRandom::maternGpuParam}.
<<setup paramsBatch, eval=TRUE,echo=FALSE>>=
myParamsBatch = 
rbind(c(shape=1.25, range=0.5, variance = 1.5, nugget = 0,anisoRatio = 1, anisoAngleRadians = 0), 
c(shape=2.15, range=0.25, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=0.55, range=1.5, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = pi/7),
c(shape=2.15, range=0.5, variance = 2, nugget = 0, anisoRatio = 4, anisoAngleRadians = -pi/7),
c(shape=2.15, range=0.5, variance = 2, nugget = 0, anisoRatio = 2, anisoAngleRadians = pi/4))
@
<<show parameterBatch,eval=TRUE>>=
myParamsBatch
paramsGpu = gpuRandom::maternGpuParam(myParamsBatch, type=theType)
@
Step 3, compute Mat\'ern covariance matrices using \fct{gpuRandom::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4800 \times 4800$ and are stacked by row in the output \code{maternCov}. 
<<r outputBatchCreate, eval=TRUE>>=
maternCov = vclMatrix(0, nrow(paramsGpu)*nrow(coordsGpu), 
                         nrow(coordsGpu),type=theType)
dim(maternCov)
gpuRandom::maternBatch(maternCov, coordsGpu, paramsGpu,
                       Nglobal=c(128,64), Nlocal=c(16,4))
@
Step 4, the first argument \code{maternCov} in \fct{gpuRandom::cholBatch} specifies the matrix object to take Cholesky decomposition. Unit lower triangular matrices $L_i$'s are returned and stacked by row in \code{maternCov}, \code{diagMat} stores the diagonal values of each $D_i$ in a row, for example, if each batch $\Sigma_i$ is of size $n \times n$, then each batch $L_i$ is $n \times n$, and each batch $D_i$ is $1 \times n$.
<<r cholBach, eval=TRUE>>=
diagMat = vclMatrix(0, nrow(paramsGpu), ncol(maternCov), type = theType)
gpuRandom::cholBatch(maternCov, diagMat, numbatchD=nrow(myParamsBatch),
                     Nglobal= c(512, 8),
                     Nlocal= c(32, 8),
                     NlocalCache=2000)
@
% \begin{gather}
% \begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
% \end{bmatrix}
%  \rightarrow
%  \begin{bmatrix}
%   L_{1} \\ L_{2} \\L_{3} \\ \vdots
%   \end{bmatrix} \text{and}
%   \begin{bmatrix}
%   D_{1} \\ D_{2} \\D_{3} \\ \vdots
%   \end{bmatrix}
% \end{gather}
Step 5, generate 2 standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{gpuRandom::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
<<r randomNormalsSim1, eval=TRUE, cache=FALSE>>=
streamsGpu <- gpuRandom::CreateStreamsGpu(Nstreams=128*64, keepInitial=TRUE)
@
<<r randomNormalsSim2, eval=TRUE, cache=FALSE>>=
zmatGpu = gpuRandom::rnorm(c(nrow(maternCov),2), 
                           streams=streamsGpu, Nglobal=c(128,64),
                           type = theType)
@
Step 6, use \fct{gpuRandom::multiplyLowerDiagonalBatch} to compute $U = L * D^{(1/2)}* Z$ in batches, see the following illustration. \code{simMat} is the output matrix for $U$, \code{maternCov}, \code{diagMat}, and \code{zmatGpu} correspond to the matrices of $L$, $D$ and $Z=(Z_1, Z_2)$ respectively.
<<r multLowerdiag_LDZ, eval=TRUE, cache=FALSE, results='hide'>>=
simMat = vclMatrix(0, nrow(zmatGpu), ncol(zmatGpu), 
                   type = theType)

gpuRandom::multiplyLowerDiagonalBatch(
  simMat, maternCov, diagMat, zmatGpu,
  diagIsOne = 1L, # diagonal of L is one
  transformD = "sqrt", # take the square root of each element of D
  Nglobal,
  Nlocal,
  NlocalCache)
@
\begin{gather}
 \begin{bmatrix}  L_{1} \\ L_{2} \\L_{3} \\ \vdots
 \end{bmatrix} 
 *
  \begin{bmatrix}
   D_{1} \\ D_{2} \\D_{3} \\ \vdots
   \end{bmatrix} 
   *
   \begin{bmatrix}
   Z_{11} & Z_{12}
   \end{bmatrix}
  =
 \begin{bmatrix}
   L_1D_1Z_{11} & L_1D_1Z_{12} \\
   L_2D_2Z_{11} & L_2D_2Z_{12} \\
   L_3D_3Z_{11} & L_3D_3Z_{12} 
  \end{bmatrix}
\end{gather}
Step 7, Finally, plot the 10 simulated Gaussian random surfaces.
<<r plotSimSetup, echo=TRUE, eval=TRUE, cache=FALSE>>=
library(raster)
simRaster = brick(myRaster, nl = ncol(simMat)*nrow(paramsGpu))
values(simRaster) = as.vector(as.matrix(simMat))
names(simRaster) = apply(expand.grid('par',1:nrow(paramsGpu), 
                                     'sim', 1:ncol(simMat)), 1, paste, collapse='')
@
\begin{figure}[H]
\centering
<<maternplot, echo=TRUE, fig.show='hold', eval=TRUE, dev='pdf',cache=FALSE>>=
par(mar=rep(0.1, 4))
for(D in names(simRaster)) {
  plot(extent(simRaster))
  plot(simRaster[[D]], legend=FALSE, add=TRUE)
}
@
\caption{\label{fig5} }
\end{figure}
